### 2023-02-04 chapter0/chapter0.tex
Removed commas after i.e.

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements. However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules. However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, \textit{i.e.}, DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

@@

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions. In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.}, eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat. Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.

>>>>

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements. However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules. However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, \textit{i.e.} DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

@@

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions. In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat. Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.
